Short paper : Automatic predicate testing in formal certification

### 2016-04-27 ProofsAndTests/paper-TAP/content/1-introduction.tex
continuing...
almost ready for the camera-ready version

One way to increase the confidence in the software we build is to formally prove its correctness using a proof assistant. Proofs assistants enable to write code, logical statements and proofs in the same language, and offer the guarantee that every proof will be automatically checked. Many of them are functional programming languages, like Coq, Idris and Agda, and others, like the B-Toolkit belong to the imperative paradigm. These different paradigms are internally supported by different logic. Systems like Coq, Idris and Agda are based on various higher order logics (CoC, a variant of ML and LUO respectively) and are realisations of the Curry-Howard correspondence, while the formal B method is based on Hoare logic. These different foundations lead to different philosophies and different ways to implement and verify a software, but all of them greatly increase the confidence of the produced software. However, these guarantees tend to be too often considered as perfect, when they are in fact far from it. Knuth was --certainly ironically-- saying ``Beware of bugs in the above code; I have only proved it correct, not tried it". The reality is precisely that a proof isn't enough. When we prove the correctness of a function, we only gain the guarantee expressed by the proven lemma, and nothing more.

@@

\begin{lstlisting}
data isSorted : {T:Type} -> (Tord : Order T) -> (l:List T) -> Type where
    NilIsSorted : (Tord : Order T) -> isSorted Tord []
    SingletonIsSorted : (Tord : Order T) -> (x:T) -> isSorted Tord [x]
    ConsSorted : {Tord : Order T} -> (h1:T) -> (h2:T) -> (t:List T) 
                 -> (isSorted Tord (h2::t)) -> (h1 $\leq$ h2) 
                 -> (isSorted Tord (h1::(h2::t)))
\end{lstlisting}
The first and second constructor of this predicate say that $[\ ]$ and $[x]$ are sorted according to any order, and for any $x$. The third one says that a list of two or more elements is sorted if $h1 \le h2$, and if the list deprived from its head is also sorted. In order to express that the result of $sort$ is sorted, we can prove the following lemma:
$sort\_correct : \forall\ (T:Type)\ (Tord:Order\ T)\ (l:List\ T),\ isSorted\ Tord\ (sort\ l)$. The problem with this specification is that it doesn't says anything about the content of the output. The function $sort$ could just return the empty list $[\ ]$ all the time, it would still be possible to prove this correctness lemma. Here, the problem is that the function is underspecified, and it is therefore possible to write a senseless implementation, which is unfortunately provably correct. Only a careful reader could realise that the lemma $sort\_correct$ forgets to mention that the input and output list should be in bijection, meaning that everything which was originally in the input list should still be in the output, and that nothing else should have been added.

Things more nasty can happen. Imagine that the third constructor of $isSorted$, called $consSorted$ would have been written with a typo, and that the condition $(h1 \leq h2) $ would have been incorrectly written as $(h1 \leq h1)$. Any list would be seen as ``sorted", just because of this single typo, and the algorithm could for example return its input unchanged. One could object that when doing the proof of correctness, we should realize that the proof is being done too easily, without having to use the essential property that the output is being built such as any element in the list is always lower or equal than its next element. The reality is quite different because many effort are going in the direction of proof automation, which aims to let the machine generate automatically the proof for some kind of goals. For example, Coq has already a Ring prover~\cite{coq2005} and many others automations, and Idris has been recently equipped with a hierarchy of provers for algebraics structures~\cite{Slama2016}. There are even extensions to languages, such as Ltac~\cite{DelahayeLTac} and Mtac~\cite{Ziliani13} that aim to help the automation of tactics. The problem is that the machine is never going to find a proof ``too easy", and will never report that something seems weird with the specification given by the user.

>>>>

One way to increase the confidence in the software we build is to formally prove its correctness using a proof assistant. Proofs assistants enable to write code, logical statements and proofs in the same language, and offer the guarantee that every proof will be automatically checked. Many of them are functional programming languages, like Coq, Idris and Agda, and others, like the B-Toolkit belong to the imperative paradigm. These different paradigms are internally supported by different logic. Systems like Coq, Idris and Agda are based on various higher order logics (CoC, a variant of ML and LUO respectively) and are realisations of the Curry-Howard correspondence, while the formal B method is based on Hoare logic. These different foundations lead to different philosophies and different ways to implement and verify a software, but all of them greatly increase the confidence of the produced software. However, these guarantees tend to be too often considered as perfect, when they are in fact far from it. Knuth was --certainly ironically-- saying ``Beware of bugs in the above code; I have only proved it correct, not tried it". The reality is precisely that a proof is not enough. When we prove the correctness of a function, we only gain the guarantee expressed by the proven lemma, and nothing more.

@@

\begin{lstlisting}
data isSorted : {T:Type} -> (Order T) -> (List T) -> Type where
    NilIsSorted : (Tord : Order T) -> isSorted Tord []
    SingletonIsSorted : (Tord : Order T) -> (x:T) -> isSorted Tord [x]
    ConsSorted : {Tord : Order T} -> (h1:T) -> (h2:T) -> (t:List T) 
                 -> (isSorted Tord (h2::t)) -> (h1 $\leq$ h2) 
                 -> (isSorted Tord (h1::(h2::t)))
\end{lstlisting}
The first and second constructor of this predicate say that $[\ ]$ and $[x]$ are sorted according to any order, and for any $x$. The third one says that a list of two or more elements is sorted if $h1 \le h2$, and if the list deprived from its head is also sorted. In order to express that the result of $sort$ is sorted, we can prove the following lemma:
$sort\_correct : \forall\ (T:Type)\ (Tord:Order\ T)\ (l:List\ T),\ isSorted\ Tord\ (sort\ l)$. The problem with this specification is that it does not say anything about the content of the output. The function $sort$ could just return the empty list $[\ ]$ all the time, it would still be possible to prove this correctness lemma. Here, the problem is that the function is underspecified, and it is therefore possible to write a senseless implementation, which is unfortunately provably correct. Only a careful reader could realise that the lemma $sort\_correct$ forgets to mention that the input and output list should be in bijection, meaning that everything which was originally in the input list should still be in the output, and that nothing else should have been added.

Another bug in the specification could have been to simply forget the third constructor $ConsSorted$. But things more nasty can happen. Imagine that the third constructor of $isSorted$, called $consSorted$ would have been written with a typo, and that the condition $(h1 \leq h2) $ would have been incorrectly written as $(h1 \leq h1)$. Any list would be seen as ``sorted", just because of this single typo, and the algorithm could for example return its input unchanged. One could object that when doing the proof of correctness, we should realize that the proof is being done too easily, without having to use the essential property that the output is being built such as any element in the list is always lower or equal than its next element. The reality is quite different because many effort are going in the direction of proof automation, which aims to let the machine generate automatically the proof for some kind of goals. For example, Coq has already a Ring prover~\cite{coq2005} and many others automations, and Idris has been recently equipped with a hierarchy of provers for algebraics structures~\cite{Slama2016}. There are even extensions to languages, such as Ltac~\cite{DelahayeLTac} and Mtac~\cite{Ziliani13} that aim to help the automation of tactics. The problem is that the machine is never going to find a proof ``too easy", and will never report that something seems weird with the specification given by the user.

### 2016-04-27 ProofsAndTests/paper-TAP/content/2-specCanGoWrong.tex
almost ready for the camera-ready version

In order to gain confidence in the formal specifications we write, another traditional approach is to test the predicate on some values. That consists in defining a few terms, usually by hand, for which we know if the predicate should hold or not, and to prove that the predicate effectively holds when it should, and that it doesn't when it shouldn't. For example, with the predicate $isSorted$ defined above, we can prove that it holds on the list [3, 5, 7] that we know sorted.

@@

\begin{lstlisting}
decideIsSorted : (Tord : Order T) -> (l:List T) -> Dec(isSorted Tord l)
decideIsSorted Tord [] = Yes (NilIsSorted Tord)
decideIsSorted Tord [x] = Yes (SingletonIsSorted Tord x)
decideIsSorted Tord (h1::(h2::t)) with (lowerEqDec Tord h1 h2)
  decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          with (decideIsSorted Tord (h2::t))
    decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          | (Yes h2_tail_sorted) = Yes (ConsSorted h1 h2 t h2_tail_sorted h1_lower_h2)
    decideIsSorted Tord (h1::(h2::t)) | (Yes h1_lower_h2) 
          | (No h2_tail_not_sorted) = No [...]
  decideIsSorted Tord (h1::(h2::t)) | No h1_not_lower_h2 = No [...]
\end{lstlisting}
Now, in order to do \emph{tests by proof}, we can simply run the decision procedure.

\begin{lstlisting}
isSorted_test1' : Dec (isSorted natIsOrdered [3,5,7])
isSorted_test1' = decideIsSorted natIsOrdered [3,5,7] 
\end{lstlisting}
And if evaluate $isSorted\_test1'$, the system will answer $Yes$ and a proof of $isSorted\ [3,\ 5,\ 7]$, which means that the predicate has passed this test. With this technique, we can run semi-automatically a few tests on the predicate $isSorted$. It is semi-automatic in the sense that we still have to define by hand some terms that we know sorted and unsorted but we can let the machine produce the proof that the predicate holds or not on these specific values. This isn't bad --and this is in fact all that is done usually, when it is actually done-- but we would like to have a stronger guarantee, and not only that the predicate will coincide with our intuitive notion on a couple of tested terms.

>>>>

In order to gain confidence in the formal specifications we write, another traditional approach is to test the predicate on some values. That consists in defining a few terms, usually by hand, for which we know if the predicate should hold or not, and to prove that the predicate effectively holds when it should, and that it does not when it should not. For example, with the predicate $isSorted$ defined above, we can prove that it holds on the list [3, 5, 7] that we know sorted.

@@

\begin{lstlisting}
decideIsSorted : (Tord : Order T) -> (l:List T) 
                 -> Dec(isSorted Tord l)
decideIsSorted Tord [] = Yes (NilIsSorted Tord)
decideIsSorted Tord [x] = Yes (SingletonIsSorted Tord x)
decideIsSorted Tord (h1::(h2::t)) with (lowerEqDec Tord h1 h2)
  | (Yes h1_lower_h2) with (decideIsSorted Tord (h2::t))
      | (Yes h2_tail_sorted) = Yes 
      		(ConsSorted h1 h2 t h2_tail_sorted h1_lower_h2) 
      | (No h2_tail_not_sorted) = No [...]
  | (No h1_not_lower_h2) = No [...]
\end{lstlisting}
Now, in order to do \emph{tests by proof}, we can simply run the decision procedure.

\begin{lstlisting}
isSorted_test1' : Dec (isSorted natIsOrdered [3,5,7])
isSorted_test1' = decideIsSorted natIsOrdered [3,5,7] 
\end{lstlisting}
And if we evaluate $isSorted\_test1'$, the system will answer $Yes$ and a proof of $isSorted\ [3,\ 5,\ 7]$, which means that the predicate has passed this test. With this technique, we can run semi-automatically a few tests on the predicate $isSorted$. It is semi-automatic in the sense that we still have to define by hand some terms that we know sorted and unsorted but we can let the machine produce the proof that the predicate holds or not on these specific values. This is not bad --and this is in fact all that is done usually, when it is actually done-- but we would like to have a stronger guarantee, and not only that the predicate will coincide with our intuitive notion on a couple of tested terms.

### 2016-04-27 ProofsAndTests/paper-TAP/content/3-testingInside.tex
almost ready for the camera-ready version

$generateSortedList : (T:Type)\ \rightarrow\ (recEnu:RecursivelyEnumerable\ T) \rightarrow\ (Tord : Order\ T) \rightarrow\ (n:Nat) \rightarrow Stream\ (List\ T)$.

To do so, the type $T$ needs to be $RecursivelyEnumerable$, which means that there should exists a computational map $Nat\ \rightarrow\ Maybe\ c$ with the condition that this map is surjective, which means that all value of type $c$ should be hit at least once by the map : $map\_is\_surjective : (y:c)\ \rightarrow\ (x:Nat\ **\ (computableMap\ x\ =\ Just\ y))$.

>>>>

$generateSortedList : (T:Type)\ \rightarrow\ (recEnu:RecEnum\ T) \rightarrow\ (Tord : Order\ T) \rightarrow\ (n:Nat) \rightarrow Stream\ (List\ T)$.

To do so, the type $T$ needs to be recursively enumerable, which means that there should exists a computational map $Nat\ \rightarrow\ Maybe\ c$ with the condition that this map is surjective, which means that all value of type $c$ should be hit at least once by the map : $map\_is\_surjective : (y:c)\ \rightarrow\ (x:Nat\ **\ (computableMap\ x\ =\ Just\ y))$.

### 2016-04-27 ProofsAndTests/paper-TAP/content/4-differentKindOfProperties.tex
almost ready for the camera-ready version

\begin{lstlisting}
generated_implies_pred_holds : {T:Type} -> (recEnu:RecursivelyEnumerable T) 
        -> (Tord : Order T) -> (n:Nat) 
        -> (All (generateSortedList T recEnu Tord n) (\l => isSorted Tord l))
\end{lstlisting}
\begin{proof}
By induction on $n$.
When $n$ is zero, there is only one sorted list generated, which is the empty list, and we know that the empty list is sorted thanks to the constructor $NilIsSorted$. When n is some successor $(S\ pn)$, we know by using recursively the lemma on the smaller value $pn$ that all sorted lists of size $pn$ are sorted according to $isSorted$. Since the $Stream$ of all sorted lists of size $(S\ pn)$ has been made from the $Stream$ of all sorted lists of size $pn$ by adding to all of them --on the head position-- an element lower or equal to the current heads, we know that the property has been preserved at the higher rank.
\qed
\end{proof}

>>>>

\begin{lstlisting}
generated_implies_pred_holds : {T:Type} 
   -> (recEnu:RecEnum T) -> (Tord : Order T) -> (n:Nat) 
   -> (All (generateSortedList T recEnu Tord n) 
           (\l => isSorted Tord l))
\end{lstlisting}
\begin{proof}
By induction on $n$.
When $n$ is zero, there is only one sorted list generated, which is the empty list, and we know that the empty list is sorted thanks to the constructor $NilIsSorted$. When n is some successor $(S\ pn)$, we know by using recursively the lemma on the smaller value $pn$ that all sorted lists of size $pn$ are sorted according to $isSorted$. Since the $Stream$ of all sorted lists of size $(S\ pn)$ has been made from the $Stream$ of all sorted lists of size $pn$ by adding to all of them --on the head position-- an element lower or equal to the current heads, we know that the property has been preserved at the higher rank.
\qed
\end{proof}

### 2016-04-27 ProofsAndTests/paper-TAP/paper-TAP.tex
Finished
paper
almost ready for the camera-ready version

\begin{abstract}
The use of formal methods and proof assistants helps to increase the confidence in critical software. However, a formal proof is only a guarantee relative to a formal specification, and not necessary about the real requirements. There is always a jump when going from an informal specification to a formal specification expressed in a logical theory. Thus, proving the correctness of a piece of software always makes the implicit assumption that there is adequacy between the formalised specification --written with logical statements and predicates-- and the real requirements --often written in english--. Unfortunately, a huge part of the complexity lies precisely in the specification itself, and it is far from obvious that the formal specification says exactly and completely what it should say. Why should we trust more these predicates than the code that we've first refused to trust blindly, leading to these proofs? We show in this paper that the proving activity hasn't replaced the testing activity but has only changed the object which requires to be tested. Instead of testing code, we now need to test predicates. We present recent ideas about how to conduct these tests inside the proof assistant on a few examples, and how to automate them as far as possible.

\keywords{Formal certification, specification, predicate testing, proof assistant, tests, requirements}
\end{abstract}

@@

\bibliographystyle{splncs03}
\bibliography{bib-TAP}

>>>>

\begin{abstract}
The use of formal methods and proof assistants helps to increase the confidence in critical software. However, a formal proof is only a guarantee relative to a formal specification, and not necessary about the real requirements. There is always a jump when going from an informal specification to a formal specification expressed in a logical theory. Thus, proving the correctness of a piece of software always makes the implicit assumption that there is adequacy between the formalised specification --written with logical statements and predicates-- and the real requirements --often written in english--. Unfortunately, a huge part of the complexity lies precisely in the specification itself, and it is far from obvious that the formal specification says exactly and completely what it should say. Why should we trust more these predicates than the code that we've first refused to trust blindly, leading to these proofs? We show in this paper that the proving activity has not replaced the testing activity but has only changed the object which requires to be tested. Instead of testing code, we now need to test predicates. We present recent ideas about how to conduct these tests inside the proof assistant on a few examples, and how to automate them as far as possible.

\keywords{Formal certification, predicate testing, proof assistant}
\end{abstract}

@@

\bibliographystyle{splncs03}

\nocite{DeMillo1979}

\bibliography{bib-TAP}
